---
title: "Impact of Transmission on Miles Per Gallon rate"
author: "Fabio Scramim Rigo"
date: "November 12th, 2017"
output:
  html_document: default
  word_document: default
---

## Executive Summary

This article summarizes the results and analysis of the final project, in the context of the Linear Regression course of Data Science Specialization, offered by Johns Hopkins Public Health University. In it, Motor Trend magazine is interested to analyse the influence of the type of transmission (automatic/manual) on the fuel consumption rate of a set of cars.  

## Exploration Analysis and Model Choice

The `mtcars` dataset contains measurements of fuel consumption and other 10 aspects of design and performance made on 32 models of cars. In regards to the a subset of the data in which Motor Trend is interested in, the exploratory analysis shows that:

* The fuel consumption variable (MPG) has a real numeric range. In this study, the variable will be the outcome variable;
* The automatic/manual transmission variable (AM) is a factor of two values. In this study, the variable will be the predictor variable; 
* The *plot #1* from the *Appendix Section* shows that from exploratory plotting, automatic transmission is expected to lead to lower miles per gallon rates. 
* *Binomial modelling* and *Poisson modelling* do **not** fit this problem, the first because the outcome is not binary; this type of modelling is accurate for cases in which we need to predict a win/lose likelihood based on other measurements; the second because it is not a temporal problem. We are not evaluating a rate which depends on observation through time; 
* *Linear modelling* fits the problem, because for each factor it can be considered as a distribution around a mean with a residual. 


## Applying the Model

Applying the **linear modelling** to the problem, with intercept excluded, leads us to the following: 

```{r lm_no_intercept}
f <- factor(mtcars$am, labels = c("Automatic", "Manual"))
fit <- lm(mtcars$mpg ~ f - 1)
summary(fit)
```

From the summary of the linear fit, we confirm the distributions around the means which could be seen in the exploratory analysis plot #1. This is, the estimated "Miles per Gallon" rate for the Automatic cars is **17.147** and for Manual cars is **24.392**, which matches what is possible to see in the plot. To have a comparactive view, we may not exclude the intercept from the modelling: 

```{r lm_with_intercept}
lm(mtcars$mpg ~ f)$coefficients
```

In here, it is possible to see that cars with Manual transmission are expected to have on average **7.245** more MPG than a car with Automatic transmission. From the Stardard Error and P Value measurements it is possible to infer that the result has high confidence. Even so, let's quantify the uncertainty and making a residual analysis to confirm the assumption. 

```{r confidence}
confint(fit)
```

There is no intersection between the two confidence intervals, what means that the transmission type defines two well defined regions. Still, there are some outlier points, and we should interpret the residuals to discover the nature of those ouliers. More, the *plot #2* from the *Appendix Section* shows that residuals have low leverage (in the range **0.05 ~ 0.08**), what means, they have little power to cause noise on the calculated linear fit and are located in regions that have low impact and do not change the overall result substantially. 

## Conclusion

* **Automatic transmission leads to a worse fuel consumption rate when compared with manual transmission**;
* The fuel consumption rate for an automatic transmission car is expected to be **7.245 MpG less** than for an manual transmission car. 


## Appendix

### Plot #1: Exploratory Analysis

```{r exploratory_analysis, echo= FALSE}
plot(y = mtcars$mpg, x = factor(mtcars$am, labels = c("Automatic", "Manual")), xlab = "Transmission Type", ylab = "Miles per Gallon")
```

### Plot #2: Residuals Analysis

```{r residuals, echo= FALSE}
par(mfrow = c(2,2))
plot(fit)
par(mfrow = c(1,1))
```
